# Bottle Necks in Local Big Data Analytics


## Case study: Data Import and Memory Allocation

Consider the first steps of a data pipeline in R. The first part of our script to import and clean the data looks as follows.

```{r eval = FALSE}
###########################################################
# Big Data Statistics: Flights data import and preparation
#
# U. Matter, January 2019
###########################################################

# SET UP -----------------

# fix variables
DATA_PATH <- "data/flights.csv"

# DATA IMPORT ----------------
flights <- read.csv(DATA_PATH)

# DATA PREPARATION --------
flights <- flights[,-1:-3]


```

```{r eval=TRUE, echo=FALSE}
# fix variables
DATA_PATH <- "../../data/flights.csv"

```


When running this script, we notice that some of the steps need a certain amount of time to process. Moreover, while none of these steps obviously involves a lot of computation (such as a matrix inversion or numerical optimization), it quite likely involves memory allocation. We first read data into RAM (allocated to R by our operating system). It turns out that there are different ways to allocate RAM when reading data from a CSV file. Depending on the amount of data to be read in, one or the other approach might be faster. We first investigate the RAM allocation in R with `mem_change()` and `mem_used()`.

```{r}

# SET UP -----------------

# fix variables
DATA_PATH <- "data/flights.csv"
# load packages
library(pryr) 


# check how much memory is used by R (overall)
mem_used()

# check the change in memory due to each step

# DATA IMPORT ----------------
mem_change(flights <- read.csv(DATA_PATH))

# DATA PREPARATION --------
flights <- flights[,-1:-3]

# check how much memory is used by R now
mem_used()
```

The last result is kind of interesting. The object `flights` must have been larger right after importing it than at the end of the script. We have thrown out several variables, after all. Why does R still use that much memory? R does by default not 'clean up' memory unless it is really necessary (meaning no more memory is available). In this case, R has still way more memory available from the operating system, thus there is no need to 'collect the garbage' yet. However, we can force R to collect the garbage on the spot with `gc()`. This can be helpful to better keep track of the memory needed by an analytics script.

```{r}
gc()
```


Now, let's see how we can improve the performance of this script with regard to memory allocation. Most memory is allocated when importing the file. Obviously, any improvement of the script must still result in importing all the data. However, there are different ways to read data into RAM. `read.csv()` reads all lines of a csv file consecutively. In contrast, `data.table::fread()` first 'maps' the data file into memory and only then actually reads it in line by line. This involves an additional initial step, but the larger the file, the less relevant is this first step with regard to the total time needed to read all the data into memory. By switching on the `verbose` option, we can actually see what `fread` is doing.

```{r}
# load packages
library(data.table)

# DATA IMPORT ----------------
flights <- fread(DATA_PATH, verbose = TRUE)

```


Let's put it all together and look at the memory changes and usage. For a fair comparison, we first have to delete `flights` and collect the garbage with `gc()`.

```{r message=FALSE, warning=FALSE}

# SET UP -----------------

# fix variables
DATA_PATH <- "data/flights.csv"
# load packages
library(pryr) 
library(data.table)

# housekeeping
flights <- NULL
gc()

# check the change in memory due to each step

# DATA IMPORT ----------------
mem_change(flights <- fread(DATA_PATH))



```






## Case Study: Loops, Memory, and Vectorization

### Preparation

We first read the `economics` data set into R and extend it by duplicating its rows in order to get a slightly larger data set (this step can easily be adapted to create a very large data set).

```{r}
# read dataset into R
economics <- read.csv("data/economics.csv")
# have a look at the data
head(economics, 2)
# create a 'large' dataset out of this
for (i in 1:3) {
     economics <- rbind(economics, economics)
}
dim(economics)

```


### Na誰ve Approach (ignorant of R)

The goal of this code example is to compute the real personal consumption expenditures, assuming that `pce` in the `economics` data set provides the nominal personal consumption expenditures. Thus, we divide each value in the vector `pce` by a deflator `1.05`.

The first approach we take is based on a simple `for`-loop. In each iteration one element in `pce` is divided by the `deflator` and the resulting value is stored as a new element in the vector `pce_real`. 

```{r}
# Na誰ve approach (ignorant of R)
deflator <- 1.05 # define deflator
# iterate through each observation
pce_real <- c()
n_obs <- length(economics$pce)
for (i in 1:n_obs) {
  pce_real <- c(pce_real, economics$pce[i]/deflator)
}

# look at the result
head(pce_real, 2)


```


How long does it take?

```{r}
# Na誰ve approach (ignorant of R)
deflator <- 1.05 # define deflator
# iterate through each observation
pce_real <- list()
n_obs <- length(economics$pce)
time_elapsed <-
     system.time(
         for (i in 1:n_obs) {
              pce_real <- c(pce_real, economics$pce[i]/deflator)
})

time_elapsed

```


Assuming a linear time algorithm ($O(n)$), we need that much time for one additional row of data: 

```{r}

time_per_row <- time_elapsed[3]/n_obs
time_per_row

```


If we deal with big data, say 100 million rows, that is

```{r}
# in seconds
(time_per_row*100^4) 
# in minutes
(time_per_row*100^4)/60 
# in hours
(time_per_row*100^4)/60^2 

```



Can we improve this?


### Improvement 1: Pre-allocation of memory

In the na誰ve approach taken above, each iteration of the loop causes R to re-allocate memory because the number of elements in vector `pce_element` is changing. In simple terms, this means that R needs to execute more steps in each iteration. We can improve this with a simple trick by initiating the vector in the right size to begin with (filled with `NA` values).  

```{r}
# Improve memory allocation (still somewhat ignorant of R)
deflator <- 1.05 # define deflator
n_obs <- length(economics$pce)
# allocate memory beforehand
# Initiate the vector in the right size
pce_real <- rep(NA, n_obs)
# iterate through each observation
time_elapsed <-
     system.time(
         for (i in 1:n_obs) {
              pce_real[i] <- economics$pce[i]/deflator
})


```

Let's see if this helped to make the code faster.

```{r}

time_per_row <- time_elapsed[3]/n_obs
time_per_row

```

Again, we can extrapolate (approximately) the computation time, assuming the data set had millions of rows.

```{r}
# in seconds
(time_per_row*100^4) 
# in minutes
(time_per_row*100^4)/60 
# in hours
(time_per_row*100^4)/60^2 

```

This looks much better, but we can do even better.

### Improvement 2: Exploit vectorization

In this approach, we exploit the fact that in R 'everything is a vector' and that many of the basic R functions (such as math operators) are *vectorized*. In simple terms, this means that a vectorized operation is implemented in such a way that it can take advantage of the similarity of each of the vector's elements. That is, R only has to figure out once how to apply a given function to a vector element in order to apply it to all elements of the vector. In a simple loop, R has to go through the same 'preparatory' steps again and again in each iteration, this is time-intensive.

In this example, we specifically exploit that the division operator `/` is actually a vectorized function. Thus, the division by our `deflator` is applied to each element of `economics$pce`.

```{r}
# Do it 'the R way'
deflator <- 1.05 # define deflator
# Exploit R's vectorization!
time_elapsed <- 
     system.time(
     pce_real <- economics$pce/deflator
          )
# same result
head(pce_real, 2)

```


Now this is much faster. In fact, `system.time()` is not precise enough to capture the time elapsed...

```{r}

time_per_row <- time_elapsed[3]/n_obs

# in seconds
(time_per_row*100^4) 
# in minutes
(time_per_row*100^4)/60 
# in hours
(time_per_row*100^4)/60^2 

```




In order to measure the improvement, we use `microbenchmark::microbenchmark()` to measure the elapsed time in microseconds (millionth of a second).

```{r}
library(microbenchmark)
# measure elapsed time in microseconds (avg.)
time_elapsed <- 
  summary(microbenchmark(pce_real <- economics$pce/deflator))$mean

# per row (in sec)
time_per_row <- (time_elapsed/n_obs)/10^6

```


Now we get a more precise picture regarding the improvement due to  vectorization (again, assuming 100 million rows):

```{r}
# in seconds
(time_per_row*100^4) 
# in minutes
(time_per_row*100^4)/60 
# in hours
(time_per_row*100^4)/60^2 

```


<!-- #### What do we learn from this? -->


<!-- 1. How R allocates and deallocates memory can have a substantial effect on computation time. -->
<!--      - (Particularly, if we deal with a large data set!) -->
<!-- 2. In what way the computation is implemented can matter a lot for the time elapsed. -->
<!--      - (For example, loops vs. vectorization/apply) -->




## Case study: Bootstrapping and Parallel Processing

In this example, we estimate a simple regression model that aims to assess racial discrimination in the context of police stops.^[Note that this example aims to illustrate a point about computation in an applied econometrics context. It does not make any argument about identification or the broader research question whatsoever.] The example is based on the 'Minneapolis Police Department 2017 Stop Dataset', containing data on nearly all stops made by the Minneapolis Police Department for the year 2017.

We start with importing the data into R.
```{r message=FALSE, warning=FALSE}
url <- "https://vincentarelbundock.github.io/Rdatasets/csv/carData/MplsStops.csv"
stopdata <- data.table::fread(url) 
```

We specify a simple linear probability model that aims to test whether a person identified as 'white' is less likely to have her vehicle searched when stopped by the police. In order to take into account level-differences between different police precincts, we add precinct-indicators to the regression specification

First, let's remove observations with missing entries (`NA`) and code our main explanatory variable and the dependent variable.

```{r}
# remove incomplete obs
stopdata <- na.omit(stopdata)
# code dependent var
stopdata$vsearch <- 0
stopdata$vsearch[stopdata$vehicleSearch=="YES"] <- 1
# code explanatory var
stopdata$white <- 0
stopdata$white[stopdata$race=="White"] <- 1
```


We specify our baseline model as follows. 

```{r}
model <- vsearch ~ white + factor(policePrecinct)
```

And estimate the linear probability model via OLS (the `lm` function).

```{r}
fit <- lm(model, stopdata)
summary(fit)
```

A potential problem with this approach (and there might be many more in this simple example) is that observations stemming from different police precincts might be correlated over time. If that is the case, we likely underestimate the coefficient's standard errors. There is a standard approach to compute estimates for so-called *cluster-robust* standard errors which would take the problem of correlation over time within clusters into consideration (and deliver a more conservative estimate of the SEs). However, this approach only works well if the number of clusters in the data is roughly 50 or more. Here we only have 5. 

The alternative approach is to compute bootstrapped clustered standard errors. That is, we apply the [bootstrap resampling procedure](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) at the cluster level. Specifically, we draw $B$ samples (with replacement), estimate and record for each bootstrap-sample the coefficient vector, and then estimate $SE_{boot}$ based on the standard deviation of all respective estimated coefficient values.

```{r message=FALSE}
# load packages
library(data.table)
# set the 'seed' for random numbers (makes the example reproducible)
set.seed(2)

# set number of bootstrap iterations
B <- 10
# get selection of precincts
precincts <- unique(stopdata$policePrecinct)
# container for coefficients
boot_coefs <- matrix(NA, nrow = B, ncol = 2)
# draw bootstrap samples, estimate model for each sample
for (i in 1:B) {
     
     # draw sample of precincts (cluster level)
     precincts_i <- sample(precincts, size = 5, replace = TRUE)
     # get observations
     bs_i <- lapply(precincts_i, function(x) stopdata[stopdata$policePrecinct==x,])
     bs_i <- rbindlist(bs_i)
     
     # estimate model and record coefficients
     boot_coefs[i,] <- coef(lm(model, bs_i))[1:2] # ignore FE-coefficients
}
```


Finally, let's compute $SE_{boot}$.

```{r}
se_boot <- apply(boot_coefs, 
                 MARGIN = 2,
                 FUN = sd)
se_boot
```


Note that even with a very small $B$, computing $SE_{boot}$ takes up some time to compute. When setting $B$ to over 500, computation time will be substantial. Also note that running this code does hardly use up more memory than the very simple approach without bootstrapping (after all, in each bootstrap iteration the data set used to estimate the model is approximately the same size as the original data set). There is little we can do to improve the script's performance regarding memory. However we can tell R how to allocate CPU resources more efficiently to handle that many regression estimates.

Particularly, we can make use of the fact that most modern computing environments (such as a laptop) have CPUs with several *cores*. We can exploit this fact by instructing the computer to run the computations *in parallel* (simultaneously computing on several cores). The following code is a parallel implementation of our bootstrap procedure which does exactly that.

```{r message=FALSE}
# install.packages("doSNOW", "parallel")
# load packages for parallel processing
library(doSNOW)

# get the number of cores available
ncores <- parallel::detectCores()
# set cores for parallel processing
ctemp <- makeCluster(ncores) # 
registerDoSNOW(ctemp)


# set number of bootstrap iterations
B <- 10
# get selection of precincts
precincts <- unique(stopdata$policePrecinct)
# container for coefficients
boot_coefs <- matrix(NA, nrow = B, ncol = 2)

# bootstrapping in parallel
boot_coefs <- 
     foreach(i = 1:B, .combine = rbind, .packages="data.table") %dopar% {
          
          # draw sample of precincts (cluster level)
          precincts_i <- sample(precincts, size = 5, replace = TRUE)
          # get observations
          bs_i <- lapply(precincts_i, function(x) stopdata[stopdata$policePrecinct==x,])
          bs_i <- rbindlist(bs_i)
          
          # estimate model and record coefficients
          coef(lm(model, bs_i))[1:2] # ignore FE-coefficients
      
     }


# be a good citizen and stop the snow clusters
stopCluster(cl = ctemp)


```


As a last step, we compute again $SE_{boot}$.

```{r}
se_boot <- apply(boot_coefs, 
                 MARGIN = 2,
                 FUN = sd)
se_boot
```


### Parallelization with an EC2 instance



<!-- This short tutorial illustrates how to scale the computation of clustered standard errors shown in Lecture 3 up by running it on an AWS EC2 instance. Below we use the same source code as in the original example (see [`03_computation_memory.Rmd`](https://github.com/umatter/BigData/blob/master/materials/notes/03_computation_memory.Rmd)). Note that there are a few things that we need to keep in mind in order to make the script run on an AWS EC2 instance in RStudio Server.  -->

<!-- First, our EC2 instance is a Linux machine. Most of you are probably rather used to running R on a Mac or Windows PC. When running R on a Linux machine, there is an additional step to install R packages (at least for most of the packages): R packages need to be compiled before they can be installed. The command to install packages is exactly the same (`install.packages()`) and normally you only notice a slight difference in the output shown in the R console during installation (and the installation process takes a little longer than what you are used to). Apart from that, using R via RStudio Server in the cloud looks/feels very similar if not identical as when using R/RStudio locally. -->


Now, let's go through the bootstrap example. First, let's run the non-parallel implementation of the script. When executing the code below line-by-line, you will notice that essentially all parts of the script work exactly as on your local machine. This is one of the great advantages of running R/RStudio Server in the cloud. You can implement your entire data analysis locally (based on a small sample), test it locally, and then move it to the cloud and run it on a larger scale in exactly the same way (even with the same GUI).

```{r eval=FALSE}
# CASE STUDY: PARALLEL ---------------------------

# install packages
install.packages("data.table")
install.packages("doSNOW")

# load packages
library(data.table)


## ------------------------------------------------------------------------
stopdata <- read.csv("https://vincentarelbundock.github.io/Rdatasets/csv/carData/MplsStops.csv")

## ------------------------------------------------------------------------
# remove incomplete obs
stopdata <- na.omit(stopdata)
# code dependent var
stopdata$vsearch <- 0
stopdata$vsearch[stopdata$vehicleSearch=="YES"] <- 1
# code explanatory var
stopdata$white <- 0
stopdata$white[stopdata$race=="White"] <- 1

## ------------------------------------------------------------------------
model <- vsearch ~ white + factor(policePrecinct)

## ------------------------------------------------------------------------
fit <- lm(model, stopdata)
summary(fit)


# bootstrapping: normal approach

## ----message=FALSE-------------------------------------------------------

# set the 'seed' for random numbers (makes the example reproducible)
set.seed(2)

# set number of bootstrap iterations
B <- 50
# get selection of precincts
precincts <- unique(stopdata$policePrecinct)
# container for coefficients
boot_coefs <- matrix(NA, nrow = B, ncol = 2)
# draw bootstrap samples, estimate model for each sample
for (i in 1:B) {
  
  # draw sample of precincts (cluster level)
  precincts_i <- sample(precincts, size = 5, replace = TRUE)
  # get observations
  bs_i <- lapply(precincts_i, function(x) stopdata[stopdata$policePrecinct==x,])
  bs_i <- rbindlist(bs_i)
  
  # estimate model and record coefficients
  boot_coefs[i,] <- coef(lm(model, bs_i))[1:2] # ignore FE-coefficients
}

## ------------------------------------------------------------------------
se_boot <- apply(boot_coefs, 
                 MARGIN = 2,
                 FUN = sd)
se_boot



```


So far, we have only demonstrated that the simple implementation (non-parallel) works both locally and in the cloud. The real purpose of using an EC2 instance in this example is to make use of the fact that we can scale up our instance to have more CPU cores available for the parallel implementation of our bootstrap procedure. Recall that running the script below on our local machine will employ all cores available to an compute the bootstrap resampling in parallel on all these cores. Exactly the same thing happens when running the code below on our simple `t2.micro` instance. However this type of EC2 instance only has one core. You can check this when running the following line of code in RStudio Server (assuming the `doSNOW` package is installed and loaded):

```{r eval=FALSE}
parallel::detectCores()
```

When running the entire parallel implementation below, you will thus notice that it won't compute the bootstrap SE any faster than with the non-parallel version above. However, by simply initiating another EC2 type with more cores, we can distribute the workload across many CPU cores, using exactly the same R-script.

```{r eval=FALSE}

# bootstrapping: parallel approaach

## ----message=FALSE-------------------------------------------------------
# install.packages("doSNOW", "parallel")
# load packages for parallel processing
library(doSNOW)

# get the number of cores available
ncores <- parallel::detectCores()
# set cores for parallel processing
ctemp <- makeCluster(ncores) # 
registerDoSNOW(ctemp)


# set number of bootstrap iterations
B <- 50
# get selection of precincts
precincts <- unique(stopdata$policePrecinct)
# container for coefficients
boot_coefs <- matrix(NA, nrow = B, ncol = 2)

# bootstrapping in parallel
boot_coefs <- 
  foreach(i = 1:B, .combine = rbind, .packages="data.table") %dopar% {
    
    # draw sample of precincts (cluster level)
    precincts_i <- sample(precincts, size = 5, replace = TRUE)
    # get observations
    bs_i <- lapply(precincts_i, function(x) stopdata[stopdata$policePrecinct==x,])
    bs_i <- rbindlist(bs_i)
    
    # estimate model and record coefficients
    coef(lm(model, bs_i))[1:2] # ignore FE-coefficients
    
  }


# be a good citizen and stop the snow clusters
stopCluster(cl = ctemp)



## ------------------------------------------------------------------------
se_boot <- apply(boot_coefs, 
                 MARGIN = 2,
                 FUN = sd)
se_boot



```



## Case Study: Efficient Fixed Effects Estimation

In this case study we look into a very common computational problem in applied econometrics: estimation of a fixed effects model with various fixed-effects units (i.e., many intercepts). We look at this in the context of the study on ["Friends in High Places"](https://www.aeaweb.org/articles?id=10.1257/pol.6.3.63) by @cohen_malloy. Cohen and Malloy show that US Senators who are alumni of the same university/college tend to help each other out in votes on industrial policies if the corresponding policy is highly relevant for the state of one senator but not relevant for the state of the other senator. The data is provided along the published article and can be accessed here: [http://doi.org/10.3886/E114873V1](http://doi.org/10.3886/E114873V1). The data (and code) is provided in STATA format. We can import the main data set with the `foreign` package.

```{r message=FALSE, warning=FALSE}
# SET UP ------------------


# load packages
library(foreign)
library(data.table)
library(lmtest)

# fix vars
DATA_PATH <- "data/data_for_tables.dta"

# import data
cm <- as.data.table(read.dta(DATA_PATH))
# keep only clean obs
cm <- cm[!(is.na(yes)|is.na(pctsumyessameparty)|is.na(pctsumyessameschool)|is.na(pctsumyessamestate))] 

```


As part of this case study, we will replicate parts of Table 3 of the main article (p. 73). Specifically, we will estimate specifications (1) and (2). In both specifications, the dependent variable is an indicator `yes` that is equal to 1 if the corresponding senator voted Yes on the given bill and 0 otherwise. The main explanatory variables of interest are `pctsumyessameschool` (the percentage of senators from the same school as the corresponding senator who voted Yes on the given bill), `pctsumyessamestate` (the percentage of senators from the same state as the corresponding senator who voted Yes on the given bill), and `pctsumyessameparty` (the percentage of senators from the same party as the corresponding senator who voted Yes on the given bill) Specification 1 accounts for congress (time) fixed effects and senator (individual) fixed effects, and specification 2 account for congress-session-vote fixed effects and senator fixed effects. 

First, let us look at a very simple example to highlight where the computational burden in the estimation of such specifications is coming from. In terms of the regression model 1, the fixed effect specification means that we introduce an indicator variable (an intercept) for $N-1$ senators and $M-1$ congresses. That is, while the simple model matrix ($X$) without accounting for fixed effects has dimensions $425653\times4$. 

```{r}
# pooled model (no FE)
model0 <-   yes ~ 
  pctsumyessameschool + 
  pctsumyessamestate + 
  pctsumyessameparty 

dim(model.matrix(model0, data=cm))
```
In contrast, the model matrix of specification (1) is of dimensions $425653\times221$, and the model matrix of specification (2) even of $425653\times6929$.

```{r}
model1 <- 
  yes ~ pctsumyessameschool + pctsumyessamestate + pctsumyessameparty + 
  factor(congress) + factor(id) -1
mm1 <- model.matrix(model1, data=cm)
dim(mm1)
```

Using OLS to estimate such a model thus involves the computation of a very large matrix inversion (because $\hat{\beta}_{OLS} = (\mathbf{X}^\intercal\mathbf{X})^{-1}\mathbf{X}^{\intercal}\mathbf{y}$). In addition, the model matrix for specification 2 is about 22GB, which might further slow down the computer due to a lack of physical memory or even crash the R session altogether.

<!-- ```{r} -->
<!-- as.numeric(object.size(mm2))*9.31e-10 -->
<!-- ``` -->

In order to set a point of reference, we first estimate specification (1) with standard OLS.


```{r message=FALSE, warning=FALSE}

# fit specification (1)
runtime <- system.time(fit1 <- lm(data = cm, formula = model1))
coeftest(fit1)[2:4,]
# median amount of time needed for estimation
runtime[3]
```

As expected, this takes quite some time to compute. However, there is an alternative approach to estimating such models that substantially reduces the computational burden by "sweeping out the fixed effects dummies". In the simple case of only one fixed effect variable (e.g., only individual fixed effects), the trick is called "within transformation" or "demeaning" and is quite simple to implement. For each of the categories in the fixed effect variable the mean of the covariate and subtract the mean from the covariate's value.

```{r}
# illustration of within transformation for the senator fixed effects
cm_within <- 
  with(cm, data.table(yes = yes - ave(yes, id),
                      pctsumyessameschool = pctsumyessameschool - ave(pctsumyessameschool, id),
                      pctsumyessamestate = pctsumyessamestate - ave(pctsumyessamestate, id),
                      pctsumyessameparty = pctsumyessameparty - ave(pctsumyessameparty, id)
                      ))

# comparison of dummy fixed effects estimator and within estimator
dummy_time <- system.time(fit_dummy <- 
              lm(yes ~ pctsumyessameschool + 
                           pctsumyessamestate + pctsumyessameparty + factor(id) -1, data = cm
                         ))
within_time <- system.time(fit_within <- 
                             lm(yes ~ pctsumyessameschool + 
                           pctsumyessamestate + pctsumyessameparty -1, data = cm_within))
# computation time comparison
as.numeric(within_time[3])/as.numeric(dummy_time[3])

# comparison of estimates
coeftest(fit_dummy)[1:3,]
coeftest(fit_within)

```

Unfortunately, we cannot simply apply the same procedure in a specification with several fixed effects variables. However, @GAURE20138 provides a generalization of the linear within-estimator to several fixed effects variables. This method is implemented in the `lfe` package (@gaure_2013). With this package, we can easily estimate both fixed-effect specifications (as well as the corresponding cluster-robust standard errors) in order to replicate the original results by @cohen_malloy.

```{r warning=FALSE, message=FALSE}
library(lfe)

# model and clustered SE specifications
model1 <- yes ~ pctsumyessameschool + pctsumyessamestate + pctsumyessameparty |congress+id|0|id
model2 <- yes ~ pctsumyessameschool + pctsumyessamestate + pctsumyessameparty |congress_session_votenumber+id|0|id

# estimation
fit1 <- felm(model1, data=cm)
fit2 <- felm(model2, data=cm)
```

Finally we can display the regression table.

```{r warning=FALSE, message=FALSE}
stargazer::stargazer(fit1,fit2,
                     type="text",
                     dep.var.labels = "Vote (yes/no)",
                     covariate.labels = c("School Connected Votes",
                                          "State Votes",
                                          "Party Votes"),
                     keep.stat = c("adj.rsq", "n"))
```

